---
title: "data_descriptive"
format: html
editor: visual
---

# Data 2 - Maria

The `clean_2` data frame contains log₂ LFQ protein intensities measured in four purified mouse brain cell types: **astrocytes, microglia, neurons**, and **oligodendrocytes**. LFQ is a metric used to represent the relative abundance of each protein across the different cell-type samples.

Because the LFQ intensities in this dataset are already provided in **log₂ scale**, they can be directly compared between samples without further transformation. To focus the analysis on cell-type–specific protein expression, we keep only the first six columns of the dataset, which correspond to:

-    *Gene name*

-    *Protein name*

-    *Isolated Astrocytes*

-   *Isolated Microglia*

-    *Isolated Neurons*

-    *Isolated Oligodendrocytes*

To evaluate enrichment of each protein in a given cell type, we compute a **reference abundance** using the median log₂ LFQ intensity across all major brain samples. This provides a biologically meaningful baseline that reflects the “whole brain” proteome against which each isolated cell type can be compared. Using this reference, we calculate the **log₂ fold change** for each protein and each cell type:

*log₂ (FC cell type)​= log₂(LFQ cell type)​−median log₂(LFQ)*

This augmented dataset enables identification of **cell-type enriched proteins** and potential **marker candidates**, in line with the analyses presented in Figure 4 of the original publication.

# Data 3 - Louise

# Data 5 - Renata & Marianna

```{r}
clean_5 <- read_tsv(file = "../data/clean_data/clean_5.tsv", show_col_types = FALSE)
head(clean_5)
```

The `clean_5` data frame contains protein expression data of cultured cell types per individual replicate. As in `clean_2` , the data is expressed in log LFQ intensity, allowing us to both validate our replicates, and to determine the preferential expression of each cultured cell type.

For these goals, we are interested only in columns 1-29, while we can discard information about sequence coverage, molecular weight, and protein IDs.

# Data 17 - Angel

```{r}
head(clean_17)
```

This data frame currently has a lot of columns which describe the intensity of RNA signal and proteins in different regions of the mouse brain.

We're specially interested in the columns that describe the intensity of proteins. For this, the columns from 14 to 23 contain the right data.

For the PCA, the only needed data is the brain region and the intensity of signal, so the genes, protein names and everything can be dropped for the analysis.
